---
title: "Learn to R"
output: html_notebook
---

This document serves as an accompanying material to live BRC R programming sessions.

# Before we begin
Make sure you have R and RStudio installed and that you can launch them. Once in RStudio, go to "*Tools > Global Options...*". Under the *General* tab in the *Workspace* section untick "Restore .RData into workspace at startup" and set "Save workspace to .RData on exit" to *Never*.

# 1. What is R

**R** is a programming language developed in the 1980s and derived from an older programming language called **S**. It can be used for many different purposes, but is widely used by statisticians to clean, manipulate and analyse data. R is free and open-source. It is licensed under the [GNU project](https://www.gnu.org/home.en.html). R isn't just the language. It is also the whole R ecosystem. Over the years, many libraries have been written and shared to improve analysis workflows in R.

When running R code, the source code is inserted into the **R console** and subsequently translated into machine code by the R interpreter and executed. [RStudio](https://posit.co/download/rstudio-desktop/) is a popular **IDE** (integrated development environment) for writing R code and other documents (such as this one). It has many features that help write and debug code, but is not strictly necessary for writing R code.

# 2. Let's run some code

The "**>**" symbol in R console means that the interpreter is ready and waiting for our commands. Let's run some code.
```{r}
print("Hello World!")
```

We've just run our first command. We used the `print()` function and passed it the string "Hello World!" as an argument. The output of this command appeared immediately afterwards on the console. Great! Almost everything we run in R is a function. Now let's try something else.
```{r}
a <- 1
```
It looks like nothing happened, but when we look at the R console, we can see the prompt (>) is there and ready for more commands. This means that the command has finished. This command is very different from the previous one. Unlike `print()`, this one has no output, so the console didn't display anything. 

`<-` in the command is the assignment operator. It looks like an arrow pointing toward the left-hand side and is created by typing the lesser-than symbol followed by the minus symbol. It is an often-used combination so there is a keyboard shortcut for it in RStudio (*Alt+-*). The assignment operator tells the computer to take what is on the right-hand side and assign it to what is on the left-hand side.

We have just asked the computer to reserve a bit of memory, assign it the label `a` and then give it the value of **1**. In our case, `a` is a variable. It will now stay saved in the environment until we restart the R interpreter or until we remove `a`. It is very much like pressing *1* and then *M+* on a calculator.

Could we achieve the same with a function? The `<-` operator is actually a function as well. And it takes two arguments. The first argument is the name of the variable and the second argument is the value. Let's try this out.
```{r}
# The `<-` operator is actually a function
`<-`("a", 1)
```
This is definitely not the preferred way to use `<-`, but it achieves exactly the same thing. We had to add a few quotes to make it work, but don't worry about that for now.

# 3. Getting help
If you ever feel lost and don't know how something works, there are a few things you can do.

 + Use R online help. Use the `help(topic)` function and give it a topic to explore. It can also be accessed by using `?topic`
```{r}
# help(topic)
help("print")
# ?topic
?print
```

 + Search online. Use documentation or your favourite AI assistant.
 + Ask on forums. [Stack overflow](https://stackoverflow.com/questions) is an excellent resource.
 
### Exercise

 1. Try running `help()` on `help` itself: `help("help")`   
 1. What happens if you run just `help()`?

# 4. Classes
In R, there are several basic classes that dictate what can be done with a value. Mathematical operations can be done with numbers but not with text, for example. The 6 basic classes in R are **character, numeric, integer, logical, complex, raw**. We can find out what class a variable is by running `class(variable)`.

Variables can be converted between the different classes. This is called **casting** and is typically done by using one of the `as.class` functions.

### Character
Notice that in our very first command, we used double quotes to enclose the message. That's because we worked with text. A sequence of characters is often called a **string** in programming. In R, the string class is called **character** and is surrounded by double quotes `"` or by single quotes `'`.
```{r}
text <- "Hello"
class(text)
```
Characters can also be created by converting from other classes using `as.character()`. Note that characters that look like numbers are still characters (text) and cannot be used in mathematical operations.
```{r}
# Cast a numeric to a character
text <- as.character(1)
class(text)
# Try to do a mathematical operation with a character
text + 1
```

### Numeric
Real numbers (both whole and decimal) are represented by the **numeric** class. 
```{r}
number <- 1
class(number)
```
Variables can be converted to numeric using `as.numeric()` if such conversion makes sense. Be careful, some conversions would not make sense to a human, but are perfectly acceptable for a computer.
```{r}
# Create a character
text <- "123"
# Cast a character to a numeric
number <- as.numeric(text)
class(number)
```
### Integer
**Integer** class is used to represent whole numbers only. They can be created by using the `as.integer()` function or by appending "**L**" to a number.
```{r}
# Create an integer
int <- 1L
class(int)
# Cast a numeric to an integer
int <-  as.integer(1)
class(int)
```
**Integers** have some limitations compared to **numeric** and are generally only useful when our R code needs to interact with other programming languages or when memory efficiency is of concern.

### Logical
**Logical** class is used to represent Boolean values (**TRUE** and **FALSE**).
```{r}
# Create a logical
coding_is_great <- TRUE
class(coding_is_great)
```

### Complex
The **complex** class is used to represent complex numbers.

```{r}
# Create a complex number
complex <- 1 + 5i
class(complex)

# Creating complex numbers using the complex() function
complex <- complex(real = 1, imaginary = 5)
```

### Raw
The **raw** class represents binary data and is used for handling raw bytes.

### Missing values, NaN
**Missing values** in R are represented by the `NA` value. If a value is missing, we simply have no information about it. We do not know what value it has or even what type it is. Any operations done on `NA` therefore result in `NA`. The `is.na()` function is able to check whether a value is a missing value.
`NaN` is a short-hand for "not a number". It is the result of undefined mathematical operations, for example `0/0`. Both `is.nan()` and `is.na()` recognise `NaN` as a missing value.


### Exercise

1. Try converting various decimal numbers to integer class and check the results.
1. Try using the `is.character()` function on a variable to see if it is a character.
1. Once you have an answer, check what the class of the answer is.
1. Can you convert a numeric to a logical? What about a logical to a numeric?
1. Test if a value is `NA` with `is.na()`.
1. Compare the output of `NA == NA` to `is.na(NA)`.

# 5. Operators
An **operator** is a symbol or a function that performs operations on one or more variables or values. We have already used the assignment operator `<-` to assign a value to a variable. Operators allow us to perform arithmetic calculations, make comparisons, or apply logical operations. Below are some commonly used arithmetic, comparison and logical operators.

### Arithmetic operators
| Operator  | Action                  | Example               |
|:---------:|:-----------------------:|:---------------------:|
| `+`       | Addition                | `1 + 1`               |
| `-`       | Subtraction             | `1 - 1`               |
| `*`       | Multiplication          | `2 * 3`               |
| `/`       | Division                | `6 / 3`               |
| `^` / `**`| Exponentiation          | `2 ^ 3`               |
| `%%`      | Modulus                 | `5 %% 2`              |
| `%/%`     | Integer Division        | `5 %/% 2`             |

### Comparison operators
| Operator  | Action                  | Example               |
|:---------:|:-----------------------:|:---------------------:|
| `==`      | Equal to                | `1 == 1`, `"a" == "a"`|
| `!=`      | Not equal to            | `2 != 3`, `"a" != "b"`|
| `<`       | Less than               | `2 < 3`,  `"a" < "b"` |
| `<=`      | Less than or equal to   | `2 <= 3`, `"1" <= "2"`|
| `>`       | Greater than            | `3 > 2`,  `"5" > "2"` |
| `>=`      | Greater than or equal to| `3 >= 2`, `"a" >= "2"`|

### Logical operators
| Operator  | Action                  | Example               |
|:---------:|:-----------------------:|:---------------------:|
| `&`       | Element-wise AND        | `TRUE & TRUE`         |
| `|`       | Element-wise OR         | `TRUE | FALSE`        |
| `!`       | Logical NOT             | `!FALSE`              |
| `%in%`    | Membership              | `1 %in% c(1,2,3)`     |

### Exercise
1. Imagine R is a calculator. Use arithmetic operators with numeric variables.
1. Practice using comparison operators. What class is the answer?
1. Try using arithmetic operators on characters.
1. Use comparison operators on characters and on characters representing numbers, e.g. `"a" < "1"`.
1. Use comparison operators to compare characters to numerics. What happened? Why?

# 6. Collections
### Vectors
A **vector** is an **ordered collection** of values of the **same type**. `Character`, `numeric`, `integer` and `logical` can all form vectors. In fact, when we ran `a <- 1`, `a` is actually a numeric vector that has only one value (it's length equals 1).

A vector is created using the `c()` function. Values of a vector can then be accessed using numeric or logical vectors with the `[[]]` (for a single value only) or `[]` notation. This is called **vector indexing**, **slicing** or **subsetting**.

Operations can be done on vectors the same way they can be done with single values. Remember that single values are just a special case of a vector with only one value.

```{r}
# create a numeric vector
numeric_vector <- c(1, 2, 3, 4, 5)
# access the first value of `numeric_vector` using a numeric index
numeric_vector[[1]]
# access several values
numeric_vector[c(1, 3, 5)]
# create a character vector
character_vector <- c("a", "b", "c", "d", "e")
# access values of a character vector using a logical vector
character_vector[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
```
If the slicing vector is shorter than the sliced vector, then the shorter vector will be "recycled" to accommodate for that. **Vector recycling** also applies to some operations done on vectors.
```{r}
# vector recycling
character_vector[c(TRUE, FALSE)]
```

### List
A list is an **ordered collection** of objects of **any type**. A list is created using the `list()` function. Objects in a list can accesssed using the `[[]]` or `[]` notation, similar to vectors. However, `[]` in this case always returns a list. For extracting the actual values, `[[]]` is appropriate.

**Named lists** can be created with `list(name = value)` or by adding the names to an existing list `names(list) <- c("name1", "name2")`. Values in named lists can be accessed with `$` followed by their names.

```{r}
# create a list
a_list <- list(
  c(1,2,3,4,5),
  c("a","b","c"),
  TRUE)

# create a named list
a_list <- list(
  numbers = c(1, 2, 3, 4, 5),
  letters = c("a", "b", "c"),
  logicals = TRUE
  )

# add names to an existing list
names(a_list) <- c("numbers", "letters", "logicals")

# different ways to access objects in a list
a_list[[1]]
a_list[["letters"]]
a_list$logicals
```

### Exercise
1. Practice creating vectors of different types and accessing their values.
1. Now do the same with a list and a named list.
1. Practice different operations on vectors. For example `c(1, 2, 3) + c(3, 2, 1)`.

